Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations

### 2022-11-12 paper/paper/SPIE/article.tex
[WIP]

In the human brain, each neuron, collection of neurons, or region of interest
(ROI) is a potentially-interacting component in a highly complex system. RMT
may have potential in describing the totality of these interactions, provided
that measurements of functioning can be obtained with sufficient spatial and
temporal resolution to speak to some neurobiological or neuropsychological
phenomenon of interest.

@@

The basic insight of RMT is thus that the eigenvalues alone may provide
interesting information about highly complex systems. However, real systems are
usually noisy, and involve a mixture of random and non-random components and
interactions. RMT is statistical in nature, describing only the expected
behaviour of the spectra of \textit{iid} random matrics. To this end, a number
of summary statistics or ``spectral observables''
\cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a} can be
computed from empirically observed spectra, with these spectral observables
sometimes being better suited for further analysis, or for the comparison of
empirical observations to theory. Two such summary statistics that have been
popular\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014, matharooSpontaneousBackpainAlters2020,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016} are the \textit{spectral rigidity} and
\textit{level number variance} (``rigidity'' and ``level variance'' for short;
details in Section \ref{sect:methods}).

@@

Typical reductions include the use of a ``seed'' voxel or collection of voxels
(i.e. ROI): the mean signal over that ROI is correlated with all other N ROIs
of interest to generate a reduced functional connectivity matrix (or image)
with one correlation value at each non-seed ROI. This sort of reduction reduces
the functional connectivity to N values. This approach strongly biases the
analysis to the seed ROI.

Likewise, one can work with ROI mean signals, and work with the \(N \times N\)
matrix of these ROI correlations. This analysis [maybe cite Poldrack book for
this?] is, \textit{a priori}, sensitive to the choice of ROIs, and, in the case
of anatomical atlases / parcellations, may also lack theoretical justification.

A major disadvantage of both the above approaches, besides requiring
parcellation into ROIs, is that most such parcellation methods are atlas-based,
and thus require image registration, and, general, an entire preprocessing
pipeline to ensure that registration is successful.

RMT suggests a potentially useful reduction in the form of the eigenvalues of
the voxelwise functional connectivity. An \(N \times t\) matrix \(\bf{A}\) of
\(N\) time series of length \(t\), and where \(t \ll N\) will have a symmetric
correlation (or covariance) matrix with \(t - 1\) non-zero, positive
real-valued eigenvalues \(\Lambda\). These eigenvalues can be computed highly
efficiently via transposition of the voxelwise correlation or covariance
matrix.

@@

However, these previous studies took largely descriptive / explanatory
approaches, noting only differences in RMT across subgroups and/or conditions,
in many cases, with no formal statistical model nor formal testing of the
significance and robustness of these differences. No code was provided to allow
the reproduction of the results, and the extraction of the spectral rigidity
and level variance is non-trivial, and computationally demanding. Computation
of these spectral observables requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004},
which is, unfortunately, a poorly-documented fitting procedure that requires
making a number of difficult-to-justify decisions, and these decisions can
dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,abueleninSpectralUnfoldingChaotic2018,fossionRandommatrixSpectraTime2013,abueleninEffectUnfoldingSpectral2012,moralesImprovedUnfoldingDetrending2011}.
On top of the flexibility in the implementation and application of RMT to
empirical data, there is also the flexibility introduced by the choices in the
complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.

In addition these issues stemming from ``research degrees of freedom'' (cite
Gelman), there is also a basic baseline comparison issue: RMT is mathematically
non-trivial, the extraction of RMT spectral statistics is computationally
demanding, and eigenvalues, while easy to understand mathematically, are
difficult to interpret relative to simpler potential reductions of fMRI data.

What is missing is a rigorous, systematic, reproducible investigation of the
*predictice* value of RMT metrics across a wide variety of data, RMT analytic
choices, choice of eigenfeature, and choice of preprocessing, and for these RMT
eigenfeatures to be compared to simpler alternative baselines. The current
study aims to do just this.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36). Selection criteria was
somewhat subjective, but we required (1), that all scans have the same spatial
and temporal resolutions, (2), the the dataset comprise 10+ human subjects, and
(3), that it was reasonable to classify at the level of an entire scan, i.e.
neither task nor event timings were needed to determine class membership. This
yielded seven datasets comprising XX total subjects, and over 5000 total fMRI
volumes.

@@

The \textit{Bilinguality} dataset\cite{ds001747:1.0.0,
goldcarrieelizabethExploringRestingState2018} examined English and
Spanish-speaking monolinguals and multilinguals during a prolonged resting
state. The original study grouped participants into three subgroups: early vs
late bilinguals vs monolingual
controls\cite{goldcarrieelizabethExploringRestingState2018}. For our analysis,
we make the binary classification task be to classify \code{monolingual v
bilingual}.

@@

The \textit{Depression} dataset\cite{bezmaternykhddRestingStateClosed2021,
bezmaternykhBrainNetworksConnectivity2021} includes rs-fMRI scans frmo
non-depressed controls and mildy- or moderately-depressed subjects. The our analysis,
the binary classification task is \code{depress v control}.

@@

The \textit{Learning} dataset\cite{schapiroHumanHippocampalReplay2018,
schapiroHumanHippocampalReplay2020} has both rs-fMRI and task fMRI scans were
available for all subjects, and \code{task v rest} is chosen as the binary
classification task for our analysis.  The task is complex and difficult to
summarize here adequately, but involved multiple phases where subjects were
presented with various related abstract images, and then later tested for their
memory of certain aspects of the learned images. Thus the classification task
might also be considered learning vs. rest.

% \subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whiole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for 2 weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

% \subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT; \cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

@@

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v tas\_nonattent} classification task was constructed
using the scores of The Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioural
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

@@

We limited preprocessing methods to these steps sometimes because data required
for other typical preprocessing steps (e.g. field intensity maps, physiological
measurements), but also because we saved all pipelined intermediates to later
compare the effect of increasing degrees of pre-processing (see Section
\ref{sec:multiverse}), and it was important to limit the number of
preprocessing degrees to prevent excessive computational costs.

@@

Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity,
\(\Delta_3(L)\), is calculated for any positive real value \(L <
\max(\Lambda)\) as:
\begin{equation}
\label{eq:rigidity}
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\end{equation}
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or
equal to \(\lambda\), \(\langle \cdot \rangle_c\) denotes the average with
respect to all starting points \(c\), and where \(A\) and \(B\) denote the
slope and intercept, respectively, of the least squares fit of a straight line
to \(\eta(\lambda)\) on \([c, c+L]\)
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as a timeseries, the spectral rigidty for a value \(L\) can be is thus the
\textit{average nonlinearity} of all intervals of length \(L\) over the series.

The level number variance, \(\Sigma^2(L)\), or level variance, for short, is
closely related to the spectral rigidity \cite{mehtaRandomMatrices2004}, and is
calculated as:
\begin{equation}
\label{eq:levlvar}
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\end{equation}
where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\),
and where \(c\), \(L\), and \(\langle \cdot \rangle_c\) are as above
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as an \textit{irregular} timeseries, the level number variance is the
variation of the number of samples of all intervals of length \(L\) over the
series.

@@

\textit{Trimming Variants}. As the unfolding procedure operates on the sorted
eigenvalues, and involves fitting a smooth polynomial or exponential to these
values, extreme values are often trimmed (or omitted from fitting)
\cite{moralesImprovedUnfoldingDetrending2011,
abueleninEffectUnfoldingSpectral2012}. The number of eigenvalues to trim is
highly subjective, and not usually well-documented. In addition, with a large
number of different source matrices, it is too destructive to use the same hard
criterion (e.g. largest 10\%, largest 3) for all unfoldings.

We develop and test four trimming variants: no trimming, precision-based,
largest, and middle, with far less subjective criteria for determining trimming
thresholds. However, we relegate the details of these trimming procedures to
the Appendix \ref{sec:trimming}, or refer readers to the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated_dataset.py#L431-L444}{source
code}.

@@

When computing RMT eigenfeatures, the polynomial fitting during unfolding
has a smoothing effect, as do the additional averaging \(\langle\;\rangle_c\)
operations. However, it is possible other, simpler
transformations of the eigenvalues (e.g. uniform or Savitsky-Golay smoothing)
might have equal or greater predictive utility. We thus also test smoothed
variants of the eigenvalues as predictive features: we smooth the sorted
eigenvalues with a uniform (moving average) filter and Savitsky-Golay, using
window sizes of 3, 5, 7, and 9 to yield 8 total ``smoothed eigenvalue'' features.

@@

It could easily be the case that a combination of RMT and non-RMT eigenfeatures
are more predictive than either alone. We thus test a variety of eigenfeature
combinations, with the motivation for these particular combinations being to
test whether RMT eigenfeatures add anything beyond smoothing. Combined features
are formed by concatenation, so that if we have features \(f_1, \dots,  f_n\)
with dimensions \(p_1, \dots, p_n\), then the combined feature is \([f_1;
\dots; f_n]\) with \(\sum p_i\) dimensions.  The final combinations chosen can
be found in Table \ref{tab:features}.

@@

We might expect the largest, middle, or smallest eigenvalues to be most useful
in characterizing some system. E.g. if the smallest eigenvalues correspond to
random / noise aspects of a system, but it is differences in the nature of the
noise that most differentiate between certain systems, then here the smallest
eigenvalues may have the most predictive utility.

Likewise, the \(L\) value in each RMT eigenfeature defines the degree of
locality at which we summarize the spectrum: at small \(L\)-values, the
spectral rigidity summarizes the non-linearity of eigenvalues that are
relatively close to each other in magnitude. At large values of \(L\), the rigidity
summarizes the long-range non-linearity of the spectrum.

@@

Baseline reductions use as statistical summaries: robust measures of location
(mean, max, min); non-robust measures of location (median, \(95^{\text{th}}\)
percentile, \(5^{\text{th}}\) percentile); non-robust measures of scale
(standard deviation, range); and robust measures of scale (interquartile range,
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile). These statistics are used to summarizing / reduce the fMRI data
along the voxel dimension, yielding a feature of \(T\) dimensions. E.g. the ``mean''
feature is just the global mean signal. Exact details can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

We do not take slice variants of these baseline features, since the baseline
features are still time series. However, we do also evaluate a number of
smoothing degrees of these features, to account for noise and to be somewhat
similar to the various smoothing variants for the eigenfeatures. Each baseline
feature thus has a uniformly smoothed variant, where the size of the smoothing
window is either 1 (no smoothing), 2, 4, 8, or 16.

@@

We use a variety of stadard machine-learning classifiers available in the
Scikit-learn\cite{pedregosaScikitlearnMachineLearning2011} Python library to
solve each classification task. We use a gradient-boosted decision tree (GBDT),
random forest (RF) classifier, support vector classifier (SVC) with radial
basis function, and k-nearest neighbours classifiers with k equal to 3, 5, and
9 (KNN3, KNN5, and KNN9, respectively). We originally also attempted to test a
simple logistic regression classifier, but found that this model to frequently
failed to converge for a number of features, even after increasing iterations
significantly. In all cases the default hyperparameter values were used.

@@

\subsection{Effect of Classifier}
mAUROC distributions were generally similar across coarse or fine feature
groupings (Figure S\ref{S-fig:coarse-classifier}, S\ref{S-fig:fine-classifier}).
In the \textit{Attention} \code{task\_attend v task\_nonattend} condition, eigenfeatures were modally predictive only with
the random forest classifier, whereas in the \textit{Aging} data, a SVC was least
likely to have mAUROC \(<\) 0.5. Overall, however, differences in the mAUROC distributions
were small.

@@

\subsection{Effect of Trimming}
Trimming based on numerical precision (see Section \ref{sec:rmt-features} and
Appendix \ref{sec:trimming}), did not result in meaningfully different mAUROC
distributions in any case (Figure S\ref{S-fig:fine-trim}). However, trimming
away the largest, or both largest and smallest eigenvalues, generally had a
significant positive effect on the predictive quality of RMT features, most
especially for RMT features involving the unfolded eigenvalues. When employing
these trimming methods, these features were consistently more predictive than
not.

@@

It is likely that the unfolded eigenvalues have predictive utility because of
their smoothing and re-scaling effect. It is clear from Figure \ref{fig:unfolded},
which depicts the raw eigenvalues and unfolded eigenvalues with different trimming
procedures, that the raw eigenvalues have an extremely exponential distribution
(Figure \ref{fig:unfolded} already uses a log-scale for plotting) due to the largest
eigenvalues. The trimmed feature distributions are clearly far less pathologically distributed.
Also, as is clear from Table \ref{tab:numerical} and Figure S\ref{S-fig:fine-classifier} that
a simple combination of the raw eigenvalues plus smoothed versions of them are the most
generally-predictive features examined here.

@@

It is also difficult to quantify the practical significance of the observed mAUROC values.
With small datasets of rs- or task-fMRI data, binary classification using whole-brain
features is challenging. For subject level classification, state of the art deep learning methods
improve upon guessing by 17\% for autism \cite{bengs4DSpatioTemporalDeep2020} or 3-30\% for ADHD
\cite{riazDeepFMRIEndtoendDeep2020}, 16\% for severe depression
\cite{ramasubbuAccuracyAutomatedClassification2016}, and 23\% for obsessive compulsive disorder
\cite{takagiNeuralMarkerObsessiveCompulsive2017}. Manual feature engineering with more separable
conditions (e.g. schizophrenia) can result in classification accuracies well above 90\%
\cite{duHighClassificationAccuracy2012}, and with larger data, sophisticated custom feature
extraction methods can achieve near perfect accuracies at classifying task vs. rest
\cite{zhangCharacterizingDifferentiatingTaskbased2016}. However, for functional connectivity data
and with classical machine learning algorithms (such as SVM) we in general only expect large
prediction accuracies (e.g. greater than 80\%) when the group functional connectivities are already
strongly separated (e.g. Cohen’s \(d > 1.0\); \citenum{dansereauStatisticalPowerPrediction2017}).

In this study, the median accuracy improvement upon guessing was 0 for all features, although
eigenfeatures could improve upon guessing by 16.2\% - 33.7\% (\ref{S-tab:acc-numerical}).

>>>>

In the human brain, each neuron, collection of neurons, or region of interest
(ROI) is a potentially-interacting component in a complex system. RMT
may have potential in describing the totality of these interactions, provided
that measurements of functioning can be obtained with sufficient spatial and
temporal resolution to speak to some neurobiological or neuropsychological
phenomenon of interest.

@@

The basic insight of RMT is thus that the eigenvalues alone may provide
interesting information about highly complex systems. However, real systems are
usually noisy, and involve a mixture of random and non-random components and
interactions. RMT is statistical in nature, describing only the expected
behaviour of the spectra of \textit{iid} random matrics. To this end, a number
of summary statistics or ``spectral observables''
\cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a} can be
computed from empirically observed spectra, with these spectral observables
sometimes being better suited for further analysis, or for the comparison of
empirical observations to theory. Two such summary statistics that have been
popular\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014, matharooSpontaneousBackpainAlters2020,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016} are the \textit{spectral rigidity} and
\textit{level number variance} (``rigidity'' and ``level variance'' for short;
details in Section \ref{sec:methods}).

@@

Typical reductions include the use of a ``seed'' voxel or collection of voxels
(see e.g. \citenum{joelRelationshipSeedbasedICAbased2011,
smithCharacterizingIndividualDifferences2014}): the mean signal over that ROI
is correlated with all other N ROIs of interest to generate a reduced
functional connectivity matrix (or image) with one correlation value at each
non-seed ROI. This sort of reduction reduces the functional connectivity to N
values, but necessarily biases the analysis to the seed ROI.

Likewise, one can work with ROI mean signal reductions, or independent
component analysis reductions\cite{joelRelationshipSeedbasedICAbased2011,
smithCharacterizingIndividualDifferences2014}, and work with the \(N \times N\)
matrix of the correlations of these reductions. This analysis is, \textit{a
priori}, sensitive to the choice of ROIs, and, in the case of anatomical
atlases / parcellations for ROIs, may also lack theoretical justification.

RMT suggests a potentially useful reduction in the form of the eigenvalues of
the voxelwise functional connectivity. An \(N \times t\) matrix \(\bf{A}\) of
\(N\) time series of length \(t\), and where \(t \ll N\) will have a symmetric
correlation (or covariance) matrix with \(t - 1\) non-zero, positive
real-valued eigenvalues. These eigenvalues can be computed highly
efficiently via transposition of the voxelwise correlation or covariance
matrix.

@@

Previous studies applying RMT to functional connectivity
data\cite{wangRandomMatrixTheory2016,
wangSpectralPropertiesTemporal2015a,matharooSpontaneousBackpainAlters2020,
guRandomMatrixTheory2020} took largely descriptive / explanatory approaches,
noting only differences in RMT across subgroups and/or conditions. For example
\citenum{wangRandomMatrixTheory2016,guRandomMatrixTheory2020} use a
Kolmogorov-Smirnov test, and \citenum{wangSpectralPropertiesTemporal2015a} uses
a t-test to provide evidence of differences in various RMT metrics. However,
statistically significant differences in a metric do not imply practical
significance or predictive utility \cite{loWhySignificantVariables2015}.

In addition, the previous papers do not provide publicly-accessible code to
reproduce results. The extraction of the spectral rigidity and level variance
is computationally-demanding and non-trivial, requiring an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004},
which is, unfortunately, a poorly-documented fitting procedure that requires
making a number of arguably subjective decisions. These fitting decisions can
dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,
abueleninSpectralUnfoldingChaotic2018, fossionRandommatrixSpectraTime2013,
abueleninEffectUnfoldingSpectral2012, moralesImprovedUnfoldingDetrending2011}.
On top of the flexibility in the implementation and application of RMT to
empirical data, there is also the flexibility introduced by the choices in the
complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.
These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b} in combination
with the absence of clear code, and exclusively descriptive/explanatory
analytic approaches means that the practical utility of RMT for fMRI is largely
uninvestigated.

What is missing is a rigorous, systematic, reproducible investigation of the
predictive value of RMT metrics across a wide variety of data, analytic
choices, RMT features, and preprocessing decisions, with RMT features being
compared to simpler alternative baseline predictors. The current study aims to
remedy this.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36). Selection criteria was
somewhat subjective, but we required (1), that all scans have the same spatial
and temporal resolutions, (2), the the dataset comprise 10+ human subjects, and
(3), that it was reasonable to classify at the level of an entire scan, i.e.
neither task nor event timings were needed to determine class membership. This
yielded seven datasets.

@@

The \textit{Bilinguality} dataset\cite{ds001747:1.0.0,
goldcarrieelizabethExploringRestingState2018} examined English and
Spanish-speaking monolinguals and multilinguals during a prolonged resting
state. The original study grouped participants into three subgroups: early vs
late bilinguals vs monolingual
controls\cite{goldcarrieelizabethExploringRestingState2018}. For our analysis,
we make the binary classification task \code{monolingual v
bilingual}.

@@

The \textit{Depression} dataset\cite{bezmaternykhddRestingStateClosed2021,
bezmaternykhBrainNetworksConnectivity2021} includes rs-fMRI scans from
non-depressed controls and mildy- or moderately-depressed subjects. In our analysis,
the binary classification task is \code{depress v control}.

@@

The \textit{Learning} dataset\cite{schapiroHumanHippocampalReplay2018,
schapiroHumanHippocampalReplay2020} has both rs-fMRI and task fMRI scans
available for all subjects, and \code{task v rest} is chosen as the binary
classification task for our analysis.  The task is complex and difficult to
summarize here adequately, but involved multiple phases where subjects were
presented with various related abstract images, and then later tested for their
memory of certain aspects of the learned images. Thus, the classification task
might also be considered learning vs. rest.

% \subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for two weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

% \subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT\cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

@@

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v tas\_nonattend} classification task was constructed
using the scores of The Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioural
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

@@

We limited preprocessing methods to these steps sometimes because data required
for other typical preprocessing steps (e.g. field intensity maps, physiological
measurements) was missing, but also because all pipelined intermediates were saved to later
compare the effect of increasing degrees of pre-processing (see Section
\ref{sec:multiverse}), and it was important to limit the number of
preprocessing degrees to prevent excessive computational costs.

@@

Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity,
\(\Delta_3(L)\), is calculated for any positive real value \(L <
\max(\Lambda)\) as:
\begin{equation}
\label{eq:rigidity}
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\end{equation}
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or
equal to \(\lambda\), \(\langle \cdot \rangle_c\) denotes the average with
respect to all starting points \(c\), and where \(A\) and \(B\) denote the
slope and intercept, respectively, of the least squares fit of a straight line
to \(\eta(\lambda)\) on \([c, c+L]\)
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as a timeseries, the spectral rigidity for a value \(L\) is the
\textit{average nonlinearity} of all intervals of length \(L\) over the series.

The level number variance, \(\Sigma^2(L)\), or \textit{level variance}, for short, is
closely related to the spectral rigidity \cite{mehtaRandomMatrices2004}, and is
calculated as:
\begin{equation}
\label{eq:levlvar}
\Sigma^2(L) = \left\langle \eta^2(L, c) \right\rangle_c - \left\langle \eta(L, c) \right\rangle^2_c
\end{equation}
where \(\eta(L, c)\) is the number of unfolded eigenvalues in \([c, c+ L]\),
and where \(c\), \(L\), and \(\langle \cdot \rangle_c\) are as above
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as an irregular timeseries, the level number variance is the
variation of the number of samples in all intervals of length \(L\) over the
series.

@@

\textit{Trimming Variants}. As the unfolding procedure operates on the sorted
eigenvalues, and involves fitting a smooth polynomial or exponential to these
values, extreme values are often trimmed (omitted from fitting)
\cite{moralesImprovedUnfoldingDetrending2011,
abueleninEffectUnfoldingSpectral2012}. The number of eigenvalues to trim is
highly subjective, and not usually well-documented, and unfortunately, with a
large number of different source matrices, it is too destructive to naively use
the same hard criterion (e.g. largest 10\%, largest 3) for all unfoldings.

We develop and test four trimming variants: no trimming, precision-based,
largest, and middle, with less subjective criteria for determining trimming
thresholds. The details of these trimming procedures are straightforward, and
can be found in Appendix \ref{sec:trimming}, or the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated_dataset.py#L431-L444}{source
code}.

@@

When computing RMT eigenfeatures, the polynomial fitting during unfolding has a
smoothing effect, as do the additional averaging \(\langle\;\rangle_c\)
operations. However, it is possible other, simpler transformations of the
eigenvalues (e.g. uniform or Savitsky-Golay smoothing) might have equal or
greater predictive utility. We thus also test smoothed variants of the
eigenvalues as predictive features: the sorted eigenvalues are smoothed with
either a uniform (moving average) filter or Savitsky-Golay, using window sizes
of 3, 5, 7, and 9 to yield 8 total ``smoothed eigenvalue'' features.

@@

A combination of RMT and non-RMT eigenfeatures
could be more predictive than either alone. We test this with  a variety of eigenfeature
combinations. The particular combinations are especially chosen to
test whether RMT eigenfeatures add anything beyond smoothing.

Combined features are formed by concatenation, so that if we have features
\(f_1, \dots,  f_n\) with dimensions \(p_1, \dots, p_n\), then the combined
feature is \([f_1; \dots; f_n]\) with \(\sum p_i\) dimensions.  The final
combinations chosen can be found in Table \ref{tab:features}.

@@

The largest, middle, or smallest eigenvalues could be most useful
in characterizing some system. E.g. if the smallest eigenvalues correspond to
random / noise aspects of a system, but differences in the nature of the
system noise most differentiate between systems, then the smallest
eigenvalues may have the most predictive utility.

Likewise, the \(L\) value in each RMT eigenfeature defines the degree of
locality at which we summarize the spectrum: with small \(L\)-values, the
spectral rigidity summarizes the non-linearity of eigenvalues that are
relatively close to each other in magnitude. At large values of \(L\), the rigidity
summarizes the long-range non-linearity of the spectrum.

@@

Baseline reductions use as statistical summaries: robust measures of location
(mean, max, min); non-robust measures of location (median, \(95^{\text{th}}\)
percentile, \(5^{\text{th}}\) percentile); non-robust measures of scale
(standard deviation, range); and robust measures of scale (interquartile range,
difference between \(95^{\text{th}}\) percentile and \(5^{\text{th}}\)
percentile). These statistics are used to reduce the fMRI data
along the voxel dimension, yielding a feature of \(T\) dimensions (e.g. the ``mean''
feature is just the global mean signal). Exact details can be found in the
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/enumerables.py#L51-L69}{paper
source code}.

We do not take slice variants of these baseline features, since the baseline
features are still time series. However, we do also evaluate a number of
smoothing degrees of these features, to account for noise and to be somewhat
similar to the various smoothing variants for the eigenfeatures: each baseline
feature has a uniformly smoothed variant, where the size of the smoothing
window is either 1 (no smoothing), 2, 4, 8, or 16.

@@

We use a variety of standard machine-learning classifiers available in the
Scikit-learn\cite{pedregosaScikitlearnMachineLearning2011} Python library to
solve each classification task. We use a gradient-boosted decision tree (GBDT),
random forest (RF) classifier, support vector classifier (SVC) with radial
basis function, and k-nearest neighbours classifiers with k equal to 3, 5, and
9 (KNN3, KNN5, and KNN9, respectively). In all cases the default hyperparameter
values were used. We originally also attempted to test a simple logistic
regression classifier, but found that this model frequently failed to converge
for a number of features, even after increasing iterations significantly.

@@

\subsection{Effect of Classifier}
mAUROC distributions were generally similar across coarse or fine feature
groupings (Figure S\ref{S-fig:coarse-classifier}, S\ref{S-fig:fine-classifier}).
In the \textit{Attention} \code{task\_attend v task\_nonattend} condition, eigenfeatures were modally predictive only with
the random forest classifier, whereas in the \textit{Aging} data, a SVC was least
likely to have mAUROC \(<\) 0.5. Overall, however, differences in the mAUROC distributions
due to classifier were small.

@@

\subsection{Effect of Trimming}
Trimming based on numerical precision (see Section \ref{sec:rmt-features} and
Appendix \ref{sec:trimming}), did not result in meaningfully different mAUROC
distributions in any case (Figure S\ref{S-fig:fine-trim}). However, trimming
away the largest, or both largest and smallest eigenvalues, generally had a
significant positive effect on the predictive quality of RMT features, most
especially for RMT features involving the unfolded eigenvalues. When employing
these trimming methods, these features were consistently more predictive than
not (Figure S\ref{S-fig:fine-trim}).

@@

It is likely that the unfolded eigenvalues have predictive utility because of
their smoothing and re-scaling effect. Figure \ref{fig:unfolded}, which depicts
the raw eigenvalues and unfolded eigenvalues with different trimming
procedures, shows how the raw eigenvalues have strongly exponential
distributions, even with logarithmic axes. This is due to the
magnitude of the largest eigenvalues, and the trimmed feature distributions in
Figure \ref{fig:unfolded} are less pathologically distributed.

Also, as is clear from Table \ref{tab:numerical} and Figure S\ref{S-fig:fine-classifier}, that
a simple combination of the raw eigenvalues plus smoothed versions of them are the most
generally-predictive features examined.

@@

It is also difficult to quantify the practical significance of the observed mAUROC values.
With small datasets of rs- or task-fMRI data, binary classification using whole-brain
features is challenging. For subject level classification, state of the art deep learning methods
improve upon guessing by 17\% for autism \cite{bengs4DSpatioTemporalDeep2020} or 3-30\% for ADHD
\cite{riazDeepFMRIEndtoendDeep2020}, 16\% for severe depression
\cite{ramasubbuAccuracyAutomatedClassification2016}, and 23\% for obsessive compulsive disorder
\cite{takagiNeuralMarkerObsessiveCompulsive2017}. Manual feature engineering with more separable
conditions (e.g. schizophrenia) can result in classification accuracies well above 90\%
\cite{duHighClassificationAccuracy2012}, and with larger data, sophisticated custom feature
extraction methods can achieve near perfect accuracies at classifying task vs. rest
\cite{zhangCharacterizingDifferentiatingTaskbased2016}. However, for functional connectivity data
and with classical machine learning algorithms (such as SVM) we in general only expect large
prediction accuracies (e.g. greater than 80\%) when the group functional connectivities are already
strongly separated (e.g. Cohen’s \(d > 1.0\); \citenum{dansereauStatisticalPowerPrediction2017}).
In this study eigenfeatures could improve upon guessing by 16.2\% - 33.7\% (\ref{S-tab:acc-numerical}).
